What Can Stack Traces Tell us about the Use of Exceptions in Java?

### 2014-05-07 android-stacks.tex
removing duplicated tables.

\subsection{General exceptions being thrown.}

From Tables~\ref{tab:toptenjava} and ~\ref{tab:toptenandroid} we can observe that,
in both Java and Android related issues, direct instances
java.lang.RuntimeException were thrown  in 10,94\% of Java repositories and
18,88\% of Android repositories respectively. Moreover, after inspecting the
top 100 most thrown exceptions (see ~\ref{tophundrend}) we could observe that
stack traces on which direct instances of java.lang.Exception and
java.lang.Error were thrown.

Throwing general exceptions is  considered a
bad practice, because the type does not carry enough information to identify the
cause of the exceptional behaviour. In such cases, developers need to relying on
exception message which can be neither complete nor precise
~\cite{gosling2000java}.

In Table ~\ref{toptenandroid} we can observe a high prevalence of
RuntimeExceptions being thrown by the Android platform (167 out of 232).
Manually inspecting the code on which RuntimeExceptions were thrown in Android
Platform we could observe that direct instances of java.lang.RuntimeException
 were thrown due to two main reasons: (i) when there was no recovery action associated
to the exceptional condition, which means that regardless the kind of exception the
only possible action was to abort the application; and when translating exceptions
signaled by application methods. Since Android platform is heavily based on
"call backs", when a application throws a runtime exception (i.e., one of the life cicle methods of an Activity) and does it, the exception scapes to compotenst of the Android platform 
which do not have contextual information to adequately handle
the exception and simply terminates the application.

We could observe that although been known as a bad practice, the signaling general exceptions in this context it was an intended design decision when no specific recovery action can be defined for the exception.

@@

%todo{discomment after submission} As we can see from Table ~ref{typeroottab},
%that most of the exceptions signaled by third-party code (e.g., Android and
%libs) were runtime exceptions. When an exception is signaled by third party
%code, it can be due to 2 reasons: the precondition associated to third party's
%code was not obeyed; or there is a coding error on third party's code. Only
%manual code inspection can reveal the reason. However, regardless the reason it
%is a good practice that all the exceptions signaled by a third part code are
%properly documented. Manually inspecting a sample of stack traces caused by
%runtime exceptions thrown by library code (approximately 10\%) we could observe
%that none of the runtime exceptions signaled by third part code were documented
%in Javadoc - this analysis did not consider the exceptions implicitly signaled
%by JVM (e.g. NullPoiterException). The undocumented runtime exceptions are a
%serious problem as it makes very difficult or even impossible for the  client
%method to protect against  undocumented runtime exceptions signaled by library
%code. As a consequence, the undocumented runtime exception may remain uncaught
%leading to  system crashes.

>>>>

\subsection{Direct instances of RuntimeExceptions been thrown.}

From Tables~\ref{tab:toptenjava} and ~\ref{tab:toptenandroid} we can observe that,
in both Java and Android related issues, direct instances
java.lang.RuntimeException were thrown  in 10,94\% of Java repositories and
18,88\% of Android repositories respectively. 
In the Android repositories, most of such exceptions  were  thrown by the Android 
platform/os (167 out of 232).

Throwing general exceptions, such as direct instances of RuntimeException, is  considered a
bad practice, because the exception type does not carry enough information to identify the
cause of the exceptional behaviour, and as a consequence developers need to relying on
the exception message which can be neither complete nor precise
~\cite{gosling2000java}.

\subsection{Undocumente runtime exceptions been thrown.}

Manualy inspecting the code of the methods throwing direct instances of RuntimeException revealed 
another and even worse bad practice:  none of the analyzed methods 
documented this exception as part of the exception interface (e.g., using @throws annotation).

Such undocumented runtime exceptions represent a serious threat to system robustness, specially
when such exceptions are thrown by reused third party code (e.g. libraries, or framework utility code).
Since it is very difficult or even impossible for the client of such thrid party code to 
protect the application against undocumented runtime exceptions, the
 undocumented runtime exception may remain uncaught and lead to  system crashes.

To better investigate the extent of this problem we classified each stack trace according to the type of the
root exception. Table~\ref{typeroottab} presents the types of root causes of all stack traces reported on 
the Android repositories.

As we can see from Table ~ref{typeroottab},  most of the exceptions signaled by third-party code
 (e.g., Android and libs) were runtime exceptions.

 Manually inspecting the library code of methods signaling runtime exceptions 
 - this analysis did not consider the exceptions implicitly signaled  by JVM (e.g. NullPoiterException) -
 we could also observe that none of the runtime exceptions signaled by them were documented.
Examples of such methos are listed on the Table X.

This result is aligned with the results of other study conducted by from Cabral
and Marcques ~\cite{} which ialso observe that runtime exceptions in .Net programs 
are most often not documented.

On the other hand, we could observe that most of the exceptions signaled by the
set of libraries reused by Android platform (i.e., org.apache.harmony,
org.w3c.dom, sun.misc, org.apache.http, org.json, org.xml, javax.) referred in
this work as libcore, were checked exceptions. It is pointed as a good practice
for libraries (see Section ~\ref{best}) because by using checked exceptions
libraries can define a precise exception interface to its clients. Such
libraries are heavily used in several projects, and such precise exception
interface may be related to the libraries maturity.

%\textbf{

% the developer does not have access to their
%source code and thus needs to rely on the library documentation about the
%runtime exceptions that should be thrown – which, more often than not, are
%neither complete nor precise (Thomas, 2002; Sacramento et al., 2006). As a
%consequence, unhandled unchecked exceptions can be seen as one of the major
%sources of bugs in current Java systems (Jo et al., 2004). }

%Manually inspecting the code signalers which RuntimeExceptions were thrown in Android
%Platform we could observe that direct instances of java.lang.RuntimeException
% were thrown due to two main reasons: (i) when there was no recovery action associated
%to the exceptional condition, which means that regardless the kind of exception the
%only possible action was to abort the application; and when translating exceptions
%signaled by application methods. Since Android platform is heavily based on
%call backs", when a application throws a runtime exception (i.e., one of the life cicle 
%methods of an Activity) and does it, the exception scapes to compotenst of the Android platform 
%which do not have contextual information to adequately handle
%the exception and simply terminates the application.

%We could observe that although been known as a bad practice, the signaling general 
%exceptions in this context it was an intended design decision when no specific recovery 
%action can be defined for the exception.

@@
